The evolution of gene expression in the term placenta of viviparous mammals

### 2016-02-21 presentation/placenta_viviparous_presentation.Rnw
add presentation

%\documentclass[english,12pt]{article}
\documentclass[doublespacing]{bmcart_dla}
\usepackage{fontspec}
\setmainfont{FreeSans}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSans}
\setmathsf{FreeSans}
\setmathtt{FreeSans}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage[letterpaper]{geometry}
\usepackage{url}
% \usepackage{fancyhdr}
\usepackage{graphicx}
% \usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
%\usepackage[plain]{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage{sfmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
% \usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
\usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
% \newenvironment{narrow}[2]{%
%   \begin{list}{}{%
%       \setlength{\topsep}{0pt}%
%       \setlength{\leftmargin}{#1}%
%       \setlength{\rightmargin}{#2}%
%       \setlength{\listparindent}{\parindent}%
%       \setlength{\itemindent}{\parindent}%
%       \setlength{\parsep}{\parskip}}%
%   \item[]}{\end{list}}
% \def\newblock{\hskip}
% \newenvironment{paperquote}{%
%   \begin{quote}%
%      \it
%   }%
%   {\end{quote}}
% \renewcommand{\textfraction}{0.15}
% \renewcommand{\topfraction}{0.85}
% \renewcommand{\bottomfraction}{0.65}
% \renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
% \newenvironment{enumerate*}%
%   {\begin{enumerate}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{enumerate}}
% \newenvironment{itemize*}%
%   {\begin{itemize}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{itemize}}
% \setcounter{tocdepth}{3}
% \oddsidemargin 0.0in 
% \textwidth 6in
% \raggedbottom
% \clubpenalty = 10000
% \widowpenalty = 10000
% set a large emergency stretch to avoid issues with overfull boxes
% \setlength{\emergencystretch}{20em}
% make the afill smaller
% \renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
% \pagestyle{fancy}

@@

\begin{document}
\raggedright

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 0)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("scales")
library("magrittr")
library("tidyr")

@@

\begin{frontmatter}
\begin{fmbox}
\dochead{Research}%
\title{The evolution of gene expression in the term placenta of viviparous mammals}%
%
\author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
\author[addressref={aff2},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}

\begin{artnotes}
\end{artnotes}

\end{fmbox}
% \twocolumn

\begin{abstractbox}
\begin{abstract}
% \section{Abstract}
  \parttitle{Background} The placenta is arguably the most
  anatomically variable organ in mammals. Understanding gene
  expression patterns among placentas from multiple species can
  illuminate the molecular basis of placenta function as well as
  potentially enrich our knowledge of the biology of obstetrical
  syndromes.

  \parttitle{Results} We measured the expression profiles of the term
  placenta of \Sexpr{star.logs[,length(unique(species))]} species of
  therian (i.e. marsupials and “placental”) mammals using RNA-Seq. We
  identified multiple genes and transcripts which are differentially
  expressed across lineages including prolactin, relaxin, and
  galectins, as well as identifying the non-housekeeping genes which
  are core to placenta function (such as components of annexin
  complexes) in all \Sexpr{star.logs[,length(unique(species))]}
  species studied.

  \parttitle{Conclusions} While there are multiple non-housekeeping
  genes which are highly expressed in all placentas examined and are
  likely to be involved in the basic functions of the mammalian
  placenta, we see incredible lineage-specific diversity in the
  expression and expansion of many genes. These lineage-specific genes
  are likely to play a role in the anatomical variability of the
  placenta and may also be implicated in important human placental
  disorders such as pre-eclampsia.

% We identified significant shifts in gene expression along specific
% evolutionary lineages using a model of character evolution. Genes that
% showed significant elevation in expression along the human lineage
% ($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
% angiogenesis, tube morphogenesis, cell migration, and regulation of
% inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
% showed exceptionally high expression in humans; these genes are
% implicated in pre-eclampsia and may indicate that the risk of
% pre-eclampsia recently emerged on the human lineage.

\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta

\end{abstract}
\end{abstractbox}
\end{frontmatter}

@@

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of invasiveness of fetal tissues into maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(\cref{tab:gestational_periods}).

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental
mammals”. The term “placental mammals” is misleading, as extant
yinothereians (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, in which the maternal and fetal blood supply is separated by
one tissue layer, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
\DLA{To discuss maternal/fetal split with derek}

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Nannospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

One explanation for the divergent paths of the placental interface is
the distinct evolutionary interplay between maternal and fetal
strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
The maternal strategy is to requisition an adequate degree of
nutrients to the fetus without risking the health of the mother, while
the fetal strategy via paternal genes is to extract as many resources
from the mother as possible. Therefore, an interplay between these
extremes can set up an intricate “arms race”, involving the imprinting
of parental alleles, expansion of gene families, and enlistment of
existing genes to expression at the maternal-fetal interface
\cite{Haig1993:Geneticconflictsinhuman}.

One symptom of this maternal-fetal conflict is its potential affect on
health
\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
addition to having an invasive hemochorial placenta, \human possess a
deeply invasive trophoblast layer that can penetrate the myometrium of
the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
Defects in placental growth are known to be associated with the
devastating condition of pre-eclampsia, which can cause hypertension,
proteinuria, and has the potential to lead to death of the mother
and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
Although cases have been documented in \textit{P. troglodytes}
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
\textit{Gorilla gorilla}
\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
far higher incidence in \human pregnancies with as much as 4\% of
pregnancies being affected in the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
  troglodytes} and \textit{G. gorilla} also have deeply invasive
trophoblasts, although the progression of invasion in great apes may
be slower than \human
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
large-scale expression studies of great ape placentas have not been
conducted. The increasing affordability of next-generation sequencing
technology has enabled the large-scale study of specific tissues from
a wide array of species to address broader phylogenetic questions in
the evolution of gene expression
\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
technology to sequence and quantify the mRNA environment of mammalian
placentas at term, including that of the \human, \textit{Pan paniscus}
(the sister species to \textit{P. troglodytes}), \textit{Ateles
  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
\cow, \elephant, \textit{Dasypus novemcinctus}, and
\textit{Monodelphis domesticus}. These species represent four major
branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
wide array of placental morphologies as well as traits such as number
and size of offspring and length of gestation
(\cref{tab:gestational_periods}). We use this data, along with
existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
differences between expression in marsupial and eutherian mammals at
term, identify the evolution of expression in important placental
genes such as hormones and galectins, identify novel lineage-specific
expression patterns, and quantify \human specific differences in gene
expression that may be related to obstetrical disorders such as
pre-eclampsia.

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

\subsection{Collection of Placental Tissue and RNA Extraction}

Fetal tissue was collected from the placenta of eight mammalian
species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

% \DLA{Add details for new sequences and fix numbers}

\subsection{Sequencing}

>>>>

\documentclass[ignorenonframetext]{beamer}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{fancybox}
\usepackage{booktabs}
% \usepackage{multirow}
\usepackage{acronym}
\usepackage{qrcode}
\usepackage{cleveref}
\usepackage[absolute,overlay]{textpos}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
\addbibresource{references.bib}

\usepackage{texshade}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usepackage{xspace}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\renewcommand*{\bibfont}{\tiny}
\usepackage{mathtools,cancel}

\renewcommand{\CancelColor}{\color{red}} %change cancel color to red

\makeatletter
\let\my@cancelto\cancelto %copy over the original cancelto command
\newcommand<>{\cancelto}[2]{\alt#3{\my@cancelto{#1}{#2}}{\mathrlap{#2}\phantom{\my@cancelto{#1}{#2}}}}
% redefine the cancelto command, using \phantom to assure that the
% result doesn't wiggle up and down with and without the arrow
\makeatother

\usepackage{multirow}
\usepackage{array}

\mode<presentation>{ 
  \usetheme{CambridgeUS}
  \usecolortheme{crane}
%   % http://identitystandards.illinois.edu/graphicstandardsmanual/generalguidelines/colors.html
  \definecolor{ilboldblue}{HTML}{002058}
  \definecolor{ilboldorange}{HTML}{E87722}
  \definecolor{ilblue}{HTML}{606EB2}
  \definecolor{ilorange}{HTML}{D45D00}
%   \setbeamercolor{alerted text}{fg=ilboldorange}
%   \setbeamercolor*{palette primary}{fg=ilboldblue,bg=ilorange!70!white}
%   \setbeamercolor*{palette secondary}{fg=ilboldblue,bg=ilorange!60!white}
%   \setbeamercolor*{palette tertiary}{bg=ilblue!60!white,fg=ilboldorange}
%   \setbeamercolor*{palette quaternary}{fg=ilboldblue,bg=ilorange!80!white}
%   
%   \setbeamercolor*{sidebar}{fg=ilorange,bg=ilboldblue}
%   
%   \setbeamercolor*{palette sidebar primary}{fg=ilblue!10!white,bg=ilorange}
%   \setbeamercolor*{palette sidebar secondary}{fg=ilorange}
%   \setbeamercolor*{palette sidebar tertiary}{fg=ilblue}
%   \setbeamercolor*{palette sidebar quaternary}{fg=ilorange}
%   
%   % \setbeamercolor*{titlelike}{parent=palette primary}
%   \setbeamercolor{titlelike}{parent=palette primary,fg=ilboldblue,bg=ilorange}
%   \setbeamercolor{frametitle}{fg=ilboldorange,bg=ilblue!20!white}
%   \setbeamercolor{frametitle right}{fg=ilboldblue,bg=ilorange}
%   
%   \setbeamercolor*{separation line}{}
%   \setbeamercolor*{fine separation line}{}
%   \setbeamercovered{transparent}  
  \logo{\begin{tikzpicture}% Pale figure
      {\node[opacity=0.734]{\IfFileExists{figures/igb_wordmark.pdf}{\includegraphics[width=2cm,height=1cm,keepaspectratio]{figures/igb_wordmark}}{}%
        };}%
    \end{tikzpicture}}
}

\title[Placenta Gene Expression]{The evolution of gene expression in the term placenta of viviparous mammals}
\author[Don L Armstrong \textit{et al.}]{Don L Armstrong, Michael R.
  McGowen, Amy Weckle, Jason Caravas, Dalen Agnew, Kurt Benirschke,
  Sue Savage-Rumbaugh, Eviatar Nevo, Chong J Kim, Günter P. Wagner,
  Roberto Romero, and Derek E. Wildman}
\institute[IGB]{Institute for Genomic Biology, Computing Genomes 
  for Reproductive Health, University of Illinois, Urbana-Champaign}
\subject{Placenta Expression}

\date{\today}
\renewcommand\cite[1]{\autocite{#1}}
% \pagestyle{fancy}

@@

\begin{document}

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.8\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 0)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("scales")
library("magrittr")
library("tidyr")

@@

\IfFileExists{figures/relevant_xkcd.png}{\frame[plain]{\centering \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{figures/relevant_xkcd.png}

    \url{https://xkcd.com/1605/}}}

\frame[plain]{\titlepage}

@@

\begin{frame}{Function and Evolution of Placenta}
\begin{itemize}
\item Placenta functions in the transfer of nutrients, oxygen, and
  waste between mother and offspring
\item Placenta is highly variable structure and form in mammals, even though major
  function is conserved
\item Placenta likely arose before the origin of therian mammals
\item Form relatively well characterized, but diversity of molecular
  environment of the placenta is only recently being understood
\item Possible role of evolutionary interplay between maternal and
  fetal strategies
\end{itemize}
\end{frame}

\begin{frame}{Impacts on Human Health?}
\begin{itemize}
\item Humans possess a deeply invasive trophoblast layer that can
  penetrate the myometrium of the uterus
  \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
\item Defects in placental growth can lead to pre-eclampsia, which can
  cause hypertension, proteinuria, and mortality of mother and/or
  infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
\item Pre-eclampsia has been seen in \textit{Pan troglodytes}
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
\textit{Gorilla gorilla}
\cite{Thornton.ea1992:Convulsionsinpregnancyin}
\item Far higher incidence in \textit{Homo sapiens}, with as many as 
  4\% of pregnancies United States affected
  \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
\end{itemize}
\end{frame}

\begin{frame}{Overall Hypotheses}
\begin{itemize}
\item Gene expression patterns correlate with placenta morphology
\item Altered expression of genes correlates with invasiveness, and
  therefore pre-eclampsia
\item Maternal/Fetal evolutionary strategies will be reflected in gene
  expression patterns
\end{itemize}
\end{frame}

\begin{frame}{Species Studied}
\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\end{figure}
\end{frame}

\begin{frame}{Variation in Gestational Periods and Litter Size}
\tiny
\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
%  \caption{Gestational periods in days and average liter sizes for studied species.}
%  \label{tab:gestational_periods}
\end{table}
\end{frame}

\begin{frame}{Variations in Placenta Morphology}
\begin{table}
\begin{tabular}{c c c}
  \toprule
  Species & Interhemal Membrane & Pattern \\
  \midrule
  \smonkey & hemochorial & trabecular \\
  \cow     & epitheliochorial & villous \\
  \dog     & endotheliochorial & labyrinthine \\
  \armadillo & hemochorial & trabecular \\
  \horse & epitheliochorial & villous \\
  \elephant & endotheliochorial & labyrinthine \\
  \human & hemochorial & villous \\
% http://onlinelibrary.wiley.com/doi/10.1002/jez.1402660511/pdf
  \opossum & epitheliochorial & choriovitelline \\
  \mouse & hemotrichorial & labyrinthine \\
  \galili & hemotrichorial & labyrinthine \\
  \sheep & epitheliochorial & villous \\
  \bonobo & hemochorial & villous \\
% http://www.biolreprod.org/content/early/2014/10/17/biolreprod.114.124818.short
  \carmeli & hemotrichorial & labyrinthine \\
  \pig & epitheliochorial & diffuse \\
  \bottomrule
\end{tabular}
\end{table}
\end{frame}

@@

\subsection{Tissue/Extraction/Sequencing}

\begin{frame}{Collection of Placental Tissue}
\begin{itemize}
\item Collected from 8 mammalian species
\item Only fetal portion was sampled
\item Sequenced using GAII with an insert size of 230 bp
\item Four additional species and additional \mouse and \human samples
  were obtained from SRA
\end{itemize}
\end{frame}
